---
title: "R Programming"
subtitle: "The basics"
author: "Serkan Korkmaz"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: lumen
    toc_float: true
---

# Introduction

In this vignette the basics of `R` is presented. 


## Variables

A variable is named `object` that holds the information of.


```{r}
# 1) Define a variables;
# Define as 1, b as 2 and c as 'R Programming'

a <- 1
b <- 2
c <- 'R Programming'
```

When a `variable` is defined, and its value is assigned we can use it as we want to. We can, for example, add `a` and `b` together without having to write out its value. 

```{r}
# 1) Add a and b
a + b
```

A variable is stored in the global `environment` until you decide to delete it - or restart `R`. The benefit of `variables` is apparent when, for example, large operations are needed. Consider the case where you want to `sum` 10.000 numbers, instead of having to add every single value you can just `sum` its variable. 

```{r}
# 1) Generate 10.000 numbers
# and store in variables
a <- runif(n = 10000)

# 2) Take the sum of the
# 10.000 numbers
sum(a)
```


### Variable types

There are different `class` of `variables`:

1. Numeric and Integers (Numbers)
2. Boolean variables (Logical values)
3. Strings (Characters)


```{r}
# 1) Define different types
# of variables
numeric_variable <- 1.5
integer_variable <- 1
boolean_variable <- TRUE
string_variable  <- 'This is a string'
```

Each `variable` serves a specific purpose in `R`, which you will discover as the course progresses. We can, for example, use a `boolean` to evaluate an `expression` conditional on the `boolean`. See the code below,

```{r}
# 1) Generate two variables
# with numeric values
a <- 1
b <- 1

# 2) Generate a variable c as the sum of
# a and b which needs to be evaluated
c <- a + b

# 3) Execute a conditional
# expression
if (c > 2) {
  
  print(
    paste(
      'The value of C IS greater than 2!'
    )
  )
  
} else {
  
  print(
    paste(
      'The value of C is NOT greater than 2!'
    )
  )
  
}


```

The example code above adds `a` and `b` and stores it in the variable `c`. The `c > 2` is an `expression` that is `TRUE` if `c > 2`. However, as the value of `c` is NOT correct, the `expression` is `FALSE` and therefore only the last part of the code is executed. 

## Data structures

In `R`, there a different data structures where information are stored; `list`, `data.frame`, `vector` and `matrix`. 

> There are, also, `arrays` but these are rarely used in social sciences. So we ignore these types of data structures.

### list

A `list` is a universal container which can hold **all** types of data. You can have a `list` of `data.frames`, `matrix` and all imaginable objects.

```{r}
# 1) Generate a list with
# two different data.frames
list_ <- list(
  mtcars,
  iris
)

# 2) Accessing each
# elements in the list 
# are done by double brackets
# which returns the element as it is
# for example
# 
# NOTE: We use head just to show
# the first couple of lines of
# the data.frame
head(
  list_[[1]]
)
```

### data.frame

A `data.frame` is the most important aspect of this class. A `data.frame` contains data of various types, it has rows and columns. Lets look at some of the inbuilt `data.frame` of `R`.

```{r}
# 1) Check MTCARS
# which is a data.frame of 
# cars
head(mtcars)
```

### vector

A `vector` is a container that typically only have **one** type of `variable`, wether its a `vector` of `boolean`, `numeric` or `character` doesn't matter. An example of a `vector` is shown below,

```{r}
# 1) Generate a vector
# of numerics. A numeric vector
numeric_vector <- c(1,2,3,4,5)

# 2) Generate a vector
# of characters. A character vector
character_vector <- c('a', 'b', 'c', 'd', 'e')

# 3) Generate a vecotr
# of booleans. A boolean vector.
boolean_vector <- c(TRUE, FALSE, FALSE, TRUE)

# 4) Print one of the
# vectors
print(
  character_vector
)
```



### matrix

A matrix is a collection of equally long `vector`. It is similar to a `data.frame`. Here is an example of a 2 by 2 matrix,

```{r}
# 1) Generate a 2x2 matrix
matrix_ <- matrix(
  data = c(
    1, 2,
    2, 1
  ),
  nrow = 2,
  ncol = 2
)

# 2) Print
# the matrix
print(
  matrix_
)
```


## Functions and loops

> The fundamental rule of coding is DRY. **DON'T REPEAT YOURSELF**

Firstly, if you repeat code your are more prone to make mistakes. Secondly, your code becomes too long to read efficiently. To follow the DRY rule we use custom `functions` and `loops`. 

Let us expore the world without custom `functions` and `loops`. Assume that you have data on the age of three individuals, you want to evaluate wether the *i*th person is older than the average. In order to do this you have to calculate the `mean` and evaluate wether `age > mean(age)`, for each person. A way to do this, is the following,

```{r}
# 1) Generate age on
# three indiciduals for the purpose of this
# example
age_vector <- c(18, 22, 17)

# 1) Evaluate the first person in
# in the vector

# 1.1) Calculate the mean age
# of the 5threepersons
mean_age <- mean(age_vector)

# 1.2) Evaluate the wether
# the person is older than
# the average by using a conditional 
# expression
if (age_vector[1] > mean_age) {
  
  print(
    paste(
      'Person 1 IS older than average'
    )
  )
  
} else {
  
  print(
    paste(
      'Person 1 is NOT older than average'
    )
  )
  
}

# 1.2) Evaluate the wether
# the person is older than
# the average by using a conditional 
# expression
if (age_vector[2] > mean_age) {
  
  print(
    paste(
      'Person 2 IS older than average'
    )
  )
  
} else {
  
  print(
    paste(
      'Person 2 is NOT older than average'
    )
  )
  
}

# 1.2) Evaluate the wether
# the person is older than
# the average by using a conditional 
# expression
if (age_vector[3] > mean_age) {
  
  print(
    paste(
      'Person 3 IS older than average'
    )
  )
  
} else {
  
  print(
    paste(
      'Person 3 is NOT older than average'
    )
  )
  
}
```


Instead of this tedious approach, we can define a `function` that evaluates the expressions for us, and use a `loop` to switch persons, so we evaluate them all. See below

```{r}
# 1) Generate age on
# three indiciduals for the purpose of this
# example
age_vector <- c(18, 22, 17)

# 2) Define a function that evaluates
# the age of the person. Call it foo.
foo <- function(
  age,
  person,
  mean_age
) {
  
  
  #' function logic;
  #' 
  #' @param age a numeric vector of length 1, to be evaluated
  #' @param person a numeric vector of length 1, which indicates
  #' the person numnber.
  #' @param mean_age a numeric vector of length 1 containing
  #' the mean average of all the persons you are evaluating.
  
  if (age > mean_age) {
    
      string <- paste(
        'Person', person, 'IS older than average'
      )
    
  } else {
    
      string <- paste(
        'Person', person, 'is NOT older than average'
      )
    
  }
  
  
  return(
    string
  )
  
}

# 3) Create a for loop using lapply
# NOTE: lapply is a simplified for loop specific to R
lapply(
  X = 1:length(age_vector),
  function(i) {
    
    foo(
      age = age_vector[i],
      person = i,
      mean_age = mean(age_vector)
    )
    
  }
)
```

Using `loops` and `functions` we saved roughly 70 lines of code, which in a bigger `project` corresponds to over 1000s lines of code if done correctly.


### Functions

When creating custom `functions` you should recognize that there is three elements.

1. `arguments` and `inputs`
2. `function` `logic`
3. `output`

In most cases your function has a main `ìnput`, which is an `expression` of some form. Mostly a `data.frame`. Along with the main `input` there is a set of auxillary `arguments` that are necessary given the problem you are trying to solve.

Given the `arguments` and goal - your `function` has to contain `code` which solves the problem at hand. And then, finally, the `function` should `return` the desired `output`.  In its most generic form, a `function` with N arguments is created like below,

```{r}
# 1) Create a generic function
# and call it foo. 
foo <- function(
    main_input,
    argument1,
    argument2,
    argument3,
    argumentN
) {
  
  #' function logic;
  #' 
  #' @param main_input an object that needs to
  #' be processen
  #' 
  #' @param argument1 an auxillary object
  #' that is needed for the function
  #' @param argument2 an auxillary object
  #' that is needed for the function 
  #' @param argument3 an auxillary object
  #' that is needed for the function 
  #' @param argumentN an auxillary object
  #' that is needed for the function 
  
  string <- paste(
    'My custom function!'
  )
  
  
  # Return statement
  # with the output you wish
  # the function returns.
  # This could also be a list, data.frame or anything
  # you want.
  return(
    string
  )
  
}
```


### For loops

You have already seen the `for loop` above, in this section we go through each element of a `for loop` using a traditional `for loop` and `lapply`. In both cases we need an `iterator` to loop through, and an object we need to evaluate.

```{r}
# Using a regular for loop

# 1) Create a numeric vector
# that we can iterate through
numeric_vector <- c(1,2,3,4,5)


# 2) Create the for loop;
for (i in 1:length(numeric_vector)) {
  
  # NOTE: i is the iterator, the 1:lenght( c(1,2,3,4,5)) is what we iterate on
  # so it goes from i = 1, i = 2, .... i = 5
  
  # Let us test if the number in the vector
  # is greater than 2.
  number <- numeric_vector[i] > 2
  
  print(
    paste(
      'The number', numeric_vector[i], 'is greater than 2'
    )
  )
  
  
}
```

Alternatively we can use `lapply` to acheive the same thing.

```{r}
# Using a regular for loop

# 1) Create a numeric vector
# that we can iterate through
numeric_vector <- c(1,2,3,4,5)


# 2) Create the for loop;
lapply(
  X = 1:length(numeric_vector),
  FUN = function(i) {
    
  # NOTE: i is the iterator, the 1:lenght( c(1,2,3,4,5)) is what we iterate on
  # so it goes from i = 1, i = 2, .... i = 5
  
  # Let us test if the number in the vector
  # is greater than 2.
  number <- numeric_vector[i] > 2
  
  paste(
      'The number', numeric_vector[i], 'is greater than 2'
    )
    
  }
)
```


In both cases we a have an `iterator` which we called `i`, and range that we iterate on `1:length(numeric_vector)` so we start from 1 and continue to 5.