---
title: "Reading Data in R"
subtitle: "The basics"
author: "Serkan Korkmaz"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: lumen
    toc_float: true
---

# Introduction

Reading `data` in `R` is the most essential part of the course. The `data` that we wil encounter is most commonly given in `csv`, `txt` or `xlsx` format. In this vignette we will discuss how to load the `data` correctly in `R`

Before we continue with the vignette, we need to have the correct `libraries` installed. So if you haven't already installed `data.table` and `readOD`, please run the following code

```{r, eval = FALSE}
# install packages
install.packages('readODS', dependencies = TRUE)
install.packages('data.table', dependencies = TRUE)
```


## Creating data

For the purpose of this vignette we will write some data to our project folder, in the specified formats above.

```{r}
# write data

# 1) CSV files

# 1.1) With semi-colon seperated
# values
data.table::fwrite(
  mtcars,
  file = 'input/mtcars_semicolon.csv',
  sep = ';'
)

# 1.2) Comma seperated
# values
data.table::fwrite(
  mtcars,
  file = 'input/mtcars_comma.csv'
)

# 2) Text files
# with empty spaces
data.table::fwrite(
  mtcars,
  file = 'input/mtcars.txt',
  sep = '\t'
)

# 3) Excel files
readODS::write_ods(
  mtcars,
  path = 'input/mtcars.xlsx'
)
```


## Reading data in R

To `read` the `data` in `R` we first need to follow the following algorithm,

1. Determine the `data format`; the extension, ie. `.csv`
2. Determine the `seperator`; the values which seperates the colums
3. Check if the data is loaded correctly

### Reading CSV files

```{r}
# 1) Read csv file
# with comma seperated values
mtcars_comma <- data.table::fread(
  'input/mtcars_comma.csv',
  sep = ","
)

# 2) Check if the data has been
# read correctly by using head
head(mtcars_comma)
```

The `data` looks correct. Lets try to `read` the comma seperated `data` with a semicolon seperator and see how `R` reacts,

```{r}
# 1) Read csv file
# with comma seperated values using
# the wrong seperator
mtcars_wrong <- data.table::fread(
  'input/mtcars_comma.csv',
  sep = ";"
)

# 2) Check if the data has been
# read correctly by using head
head(mtcars_wrong)
```

As you can see, `reading` data with the wrong separator loads all values of the data in a single column. Therefore you should always check if the `data` is read correctly.

> When using `fread` from the `data.table` library you do not have to worry about seperators, as this function automatically detects the seperator

```{r}
# 1) Read csv file
# with comma seperated values without
# specifying a seperator
mtcars_generic <- data.table::fread(
  'input/mtcars_comma.csv'
)

# 2) Check if the data has been
# read correctly by using head
head(mtcars_generic)
```

The `data` looks neat and tidy and is ready for futher analysis.

### Reading TXT files

Sometimes `data` is delivered as `.txt`, which can also be comma or semicolon seperated. But more often than not it is seperated by tabs, ie. fixed length empty space. To `read` `data` with tabular space we use `\t` as seperator.

```{r}
# 1) Read txt file
# with tab seperated values 
mtcars_generic <- data.table::fread(
  'input/mtcars.txt',
  sep = '\t'
)

# 2) Check if the data has been
# read correctly by using head
head(mtcars_generic)
```



### Reading Excel files

In rare cases you will encounter Excel files. These are a bit tricky to `read` as the `data` can be stored in different sheets. If the data is stored in the first sheet, then the `data` is easily `read` in the following way,

```{r}
# 1) Read Excel file
mtcars_excel <- readODS::read_ods(
  'input/mtcars.xlsx'
)

# 2) Check if the data has been
# read correctly by using head
head(mtcars_excel)
```

# NOTE

When you `read` data ***make sure*** that you are writing the extensions correctly. This means that if the `data` is stored as `data.xls` you `read` it as `data.xls`, not `data.xlsx`.